Estimating the replicability of psychology experiments after an initial failure to replicate

### 2024-08-06 manuscript/manuscript.Rmd
redraft discussion

Given the mix of successful and unsuccessful rescue projects, we discuss in depth a few projects where we have speculations about why they turned out the way they did.
Brief descriptions of the projects not covered in this section be found in the Appendix.

@@

In cases where re-replications failed, we do not believe any conclusions are licensed about the status of the original results.
Students conducting rescue projects put substantial effort into trying to set up rescues that had a good chance of success, but projects were constrained by budget limitations, a short timeline, and the constraint that most data collection needed to occur online.
These limitations are representative of the sort of resource limitations often faced by early-career researchers.
That said, it is possible that different results might be obtained in better-resourced settings, by scientists with more expertise and more time.

Due to the limited number of replications that are conducted and systematically reported, and the even smaller number of re-replications reported, there is a lot of uncertainty in how much replicability varies across features of the original experiment and replications.

@@

Replications are done for many reasons, and researchers often have some prior beliefs about whether a replication will succeed.
These beliefs may influence how a researcher interprets a failed replication and thus whether they think it's worth trying to rescue.
For instance, a researcher who has high credence in the main result is more likely to attribute failure to specific details in the replication and thus want to rescue by tweaking these details.
On the other side, a researcher who is dubious of the methods, results, or statistical approaches, or who has experienced failures to replicate and extend the work, may be quicker to interpret a failure as a more definitive negative result.
Interpretations may also depend on the perceived strength of the replication and what effect size is cared about.

While we focus here on a narrow type of replication and rescue, that done by an early career researcher with the goal of establishing whether this is a result *they* can build on, it exists as part of a wider ecosystem of replications that are done for different reasons with different properties.

@@

@birch2007 is about a false belief task comparing ignorance and plausible knowledge conditions.
The first replication and an additional class replication both found results directionally consistent with the original, but much smaller.
The rescue used a much larger sample size than both the original and the first replication and added an additional check to make sure the probabilities participants reported add up to 100%.
The rescue result was directionally consistent with the original and prior replications, but did not find a significant effect.

@@

@haimovitz2016 manipulated parents' mindsets with "failure-is-enhancing" or "failure-is-debilitating" primes and looked at their reactions to a hypothetical situation in which a child received a failing grade.
The first replication was directionally inconsistent with the original, but an additional class replication was consistent.
The rescue added manipulation checks which were not present in the original study or either replications.
The manipulation checks passed, but the rescue study did not recover the target effect.

@@

@jara-ettinger2022 looked at shape bias in US adults across 3 shape exemplars and found a preference to extend nouns on the basis of shape over color or material.
The replication found an above-chance preference for shape, but with very high item-level variability.
The rescuer found that the first replication used stylized versions of the original stimuli taken from a different part of the paper.
The rescue used the same images of stimuli as the original, but still found strong item-level effects and relative preferences for color and material.

@@

@porter2016 looked at how linguistic intergroup bias influences judgments of shared group membership.
The first replication was directionally against the original result.
The rescuer obtained the direct wordings of the questions from the authors of the original paper, and noticed that they were substantially different from questions in the first replication.
The rescue used the same wordings as the original paper and replicated the effect, although the manipulation check failed.

@schechtman2010 looked at how widely people generalize tones that previously have been paired with losses or gains, and found that the loss-paired tones were generalized more.
The first replication found a statistically significant effect in the opposite direction of the original study.
It also found that people took much longer to learn which tones were associated with gains vs. losses compared to the original.
The rescuer made several changes, including asking questions about the sound quality of the participant's computer.
The original study was conducted in person while the replication and rescue were online, so differences in speaker and headphone quality might have played a role in the different result.
The replication found an effect in the same direction as the 1st replication (contrary to the original study).

@@

@yeshurun2003 looked at the relationship between spatial attention and temporal resolution.
They measured the threshold where two flashes of light are percieved as one flash versus two as a function of whether the location of the flashes was pre-cued or not.
@yeshurun2003 found that cueing led to less temporal resolution.
This was independently replicated by @chica2009, but was not replicated by either the student replication or the rescue.
Both the replication and rescue used in-person participants because the study relies on precise timing and details of the screen which are difficult to control online.
The first replication had a smaller sample size than the original, while the rescue matched the original study's sample size.

>>>>

Given the mix of successful and unsuccessful rescue projects, we discuss in depth a few projects where we have speculations about why they turned out the way they did.
Brief descriptions of the projects not covered in this section can be found in the Appendix.

@@

In cases where re-replications failed, we do not believe any conclusions are licensed about the status of the original results.
The rescue projects (and first replications) were primarily conducted by first year PhD students, in a program where it is expected that PhD students author peer-reviewed papers, and where some of the rescuers already had publications. Thus, while we refer to the rescuers as students, they are investigators doing experiments that become part of the literature. 
Students conducting rescue projects put substantial effort into trying to set up rescues that had a good chance of success, but projects were constrained by budget limitations, a short timeline, and the constraint that most data collection needed to occur online. Additionally, while many students had done research before and were learning best-practices for experiments, they were not experts in the particular paradigms. These limitations are representative of the sort of knowledge and resource limitations often faced by early-career researchers. That said, it is possible that different results might be obtained in better-resourced settings, by scientists with more expertise, more time, or more privileged knowledge about how the original experiments were conducted.

Due to the limited number of replications that are conducted and systematically reported, and the even smaller number of re-replications reported, there is substantial uncertainty in how much replicability varies across features of the original experiment and replications.

@@

We have focused here on a narrow type of replication and rescue: that done by an early career researcher with the goal of establishing whether this is a result *they* can build on. We acknowledge that this type of replication exists as part of a wider ecosystem of replications that have different properties and are done for different reasons.

In many cases, researchers may have strong prior beliefs about where a replication will succeed, and these priors may influence a researchers interpretation of a failed replication, and thus their calculations about where a re-replication will succeed or is worth attempting. For instance, a researcher who has high credence in the main result is more likely to attribute failure to specific details in the replication, and thus may want to do a rescue where those details are adjusted. On the other hand, a researcher who is dubious of the methods, results, or statistical approaches in an experiment, or who has previously failed to extend the work, may be quicker to interpret a non-replication as a more definitive negative result. What a researcher chooses to do may also depend on the perceived strength of the replication, what size effects they think are relevant, and how important they think the target effect is. Thus, in some cases, when to re-replicate may be a question of prior beliefs and perceived utility rather than just likelihood of a re-replication succeeding.

@@

Write-ups of individual rescue and replication projects can be found in the repository at https://osf.io/cyk5w/.

@birch2007 is about a false belief task comparing ignorance and plausible knowledge conditions.
The first replication and an additional class replication both found results in the same direction as the original result, but much smaller. 
The rescue used a much larger sample size than both the original and the first replication and added an additional check to make sure the probabilities participants reported add up to 100%.
The rescue result was in the same direction as the original and prior replications, but did not find a significant effect.

@@

@haimovitz2016 manipulated parents' mindsets with "failure-is-enhancing" or "failure-is-debilitating" primes and looked at their reactions to a hypothetical situation in which a child received a failing grade.
The first replication found a point estimate in the opposite direction to the original effect, but an additional class replication found a directionally consistent effect.
The rescue added manipulation checks which were not present in the original study or either replications.
The manipulation checks passed, but the rescue study did not recover the target effect.

@@

@jara-ettinger2022 looked at shape bias in US adults across 3 shape exemplars and found a preference to extend nouns on the basis of shape rather than color or material.
The replication found an above-chance preference for shape, but with very high item-level variability.
The rescuer found that the first replication used stylized versions of the original stimuli taken from a different part of the paper.
The rescue used the same images of stimuli as the original, but still found strong item-level effects and relative preferences for color and material.

@@

@porter2016 looked at how linguistic intergroup bias influences judgments of shared group membership.
The first replication found an effect directionally inconsistent with the original result.
The rescuer obtained the direct wordings of the questions from the authors of the original paper, and noticed that they were substantially different from questions in the first replication.
The rescue used the same wordings as the original paper and replicated the effect, although the manipulation check failed.

@schechtman2010 looked at how widely people generalize tones that previously have been paired with losses or gains, and found that the loss-paired tones were generalized more.
The first replication found a statistically significant effect in the opposite direction of the original study.
It also found that people took much longer to learn which tones were associated with gains vs. losses compared to the original.
The rescuer made several changes, including asking questions about the sound quality of the participant's computer.
The original study was conducted in person while the replication and rescue were online, so differences in speaker and headphone quality might have played a role in the different result.
The rescue found an effect in the same direction as the 1st replication (contrary to the original study).

@@

@yeshurun2003 looked at the relationship between spatial attention and temporal resolution.
They measured the threshold where two flashes of light are perceived as one flash versus two as a function of whether the location of the flashes was pre-cued or not.
@yeshurun2003 found that cueing led to less temporal resolution.
This was independently replicated by @chica2009, but was not replicated by either the student replication or the rescue.
Both the replication and rescue used in-person participants because the study relies on precise timing and details of the screen which are difficult to control online.
The first replication had a smaller sample size than the original, while the rescue matched the original study's sample size.

### 2024-08-06 manuscript/manuscript.tex
redraft discussion

Our primary question of interest was how many of the 17 rescue projects succeeded at replicating the results in the original study.
Across the 17 rescue projects, 5 mostly or fully replicated the original results according to the subjective replication ratings.
12 got a rating of 0, 2 got a rating of .75, and 3 got a rating of 1.
Thus, a first pass answer to the question ``how often can a failed replication be salvaged?'' is 29\% (bootstrapped 95\% CI: 6\% - 53\%) of the time.

@@

Replications are done for many reasons, and researchers often have some prior beliefs about whether a replication will succeed.
These beliefs may influence how a researcher interprets a failed replication and thus whether they think it's worth trying to rescue.
For instance, a researcher who has high credence in the main result is more likely to attribute failure to specific details in the replication and thus want to rescue by tweaking these details.
On the other side, a researcher who is dubious of the methods, results, or statistical approaches, or who has experienced failures to replicate and extend the work, may be quicker to interpret a failure as a more definitive negative result.
Interpretations may also depend on the perceived strength of the replication and what effect size is cared about.

While we focus here on a narrow type of replication and rescue, that done by an early career researcher with the goal of establishing whether this is a result \emph{they} can build on, it exists as part of a wider ecosystem of replications that are done for different reasons with different properties.

@@

Birch \& Bloom (\citeproc{ref-birch2007}{2007}) is about a false belief task comparing ignorance and plausible knowledge conditions.
The first replication and an additional class replication both found results directionally consistent with the original, but much smaller.
The rescue used a much larger sample size than both the original and the first replication and added an additional check to make sure the probabilities participants reported add up to 100\%.
The rescue result was directionally consistent with the original and prior replications, but did not find a significant effect.

@@

Haimovitz \& Dweck (\citeproc{ref-haimovitz2016}{2016}) manipulated parents' mindsets with ``failure-is-enhancing'' or ``failure-is-debilitating'' primes and looked at their reactions to a hypothetical situation in which a child received a failing grade.
The first replication was directionally inconsistent with the original, but an additional class replication was consistent.
The rescue added manipulation checks which were not present in the original study or either replications.
The manipulation checks passed, but the rescue study did not recover the target effect.

@@

Jara-Ettinger et al. (\citeproc{ref-jara-ettinger2022}{2022}) looked at shape bias in US adults across 3 shape exemplars and found a preference to extend nouns on the basis of shape over color or material.
The replication found an above-chance preference for shape, but with very high item-level variability.
The rescuer found that the first replication used stylized versions of the original stimuli taken from a different part of the paper.
The rescue used the same images of stimuli as the original, but still found strong item-level effects and relative preferences for color and material.

@@

Porter et al. (\citeproc{ref-porter2016}{2016}) looked at how linguistic intergroup bias influences judgments of shared group membership.
The first replication was directionally against the original result.
The rescuer obtained the direct wordings of the questions from the authors of the original paper, and noticed that they were substantially different from questions in the first replication.
The rescue used the same wordings as the original paper and replicated the effect, although the manipulation check failed.

Schechtman et al. (\citeproc{ref-schechtman2010}{2010}) looked at how widely people generalize tones that previously have been paired with losses or gains, and found that the loss-paired tones were generalized more.
The first replication found a statistically significant effect in the opposite direction of the original study.
It also found that people took much longer to learn which tones were associated with gains vs.~losses compared to the original.
The rescuer made several changes, including asking questions about the sound quality of the participant's computer.
The original study was conducted in person while the replication and rescue were online, so differences in speaker and headphone quality might have played a role in the different result.
The replication found an effect in the same direction as the 1st replication (contrary to the original study).

@@

Yeshurun \& Levy (\citeproc{ref-yeshurun2003}{2003}) looked at the relationship between spatial attention and temporal resolution.
They measured the threshold where two flashes of light are percieved as one flash versus two as a function of whether the location of the flashes was pre-cued or not.
Yeshurun \& Levy (\citeproc{ref-yeshurun2003}{2003}) found that cueing led to less temporal resolution.
This was independently replicated by Chica \& Christie (\citeproc{ref-chica2009}{2009}), but was not replicated by either the student replication or the rescue.
Both the replication and rescue used in-person participants because the study relies on precise timing and details of the screen which are difficult to control online.
The first replication had a smaller sample size than the original, while the rescue matched the original study's sample size.

>>>>

Our primary question of interest was how many of the 17 rescue projects succeeded at replicating the results in the original study.
Across the 17 rescue projects, 5 mostly or fully replicated the original results according to the subjective replication ratings.
12 got a rating of 0, 2 got a rating of .75, and 3 got a rating of 1.
Thus, a first pass answer to the question ``how often can a failed replication be salvaged?'' is 29\% (bootstrapped 95\% CI: 12\% - 53\%) of the time.

@@

We have focused here on a narrow type of replication and rescue: that done by an early career researcher with the goal of establishing whether this is a result \emph{they} can build on. We acknowledge that this type of replication exists as part of a wider ecosystem of replications that have different properties and are done for different reasons.

In many cases, researchers may have strong prior beliefs about where a replication will succeed, and these priors may influence a researchers interpretation of a failed replication, and thus their calculations about where a re-replication will succeed or is worth attempting. For instance, a researcher who has high credence in the main result is more likely to attribute failure to specific details in the replication, and thus may want to do a rescue where those details are adjusted. On the other hand, a researcher who is dubious of the methods, results, or statistical approaches in an experiment, or who has previously failed to extend the work, may be quicker to interpret a non-replication as a more definitive negative result. What a researcher chooses to do may also depend on the perceived strength of the replication, what size effects they think are relevant, and how important they think the target effect is. Thus, in some cases, when to re-replicate may be a question of prior beliefs and perceived utility rather than just likelihood of a re-replication succeeding.

@@

Write-ups of individual rescue and replication projects can be found in the repository at \url{https://osf.io/cyk5w/}.

Birch \& Bloom (\citeproc{ref-birch2007}{2007}) is about a false belief task comparing ignorance and plausible knowledge conditions.
The first replication and an additional class replication both found results in the same direction as the original result, but much smaller.
The rescue used a much larger sample size than both the original and the first replication and added an additional check to make sure the probabilities participants reported add up to 100\%.
The rescue result was in the same direction as the original and prior replications, but did not find a significant effect.

@@

Haimovitz \& Dweck (\citeproc{ref-haimovitz2016}{2016}) manipulated parents' mindsets with ``failure-is-enhancing'' or ``failure-is-debilitating'' primes and looked at their reactions to a hypothetical situation in which a child received a failing grade.
The first replication found a point estimate in the opposite direction to the original effect, but an additional class replication found a directionally consistent effect.
The rescue added manipulation checks which were not present in the original study or either replications.
The manipulation checks passed, but the rescue study did not recover the target effect.

@@

Jara-Ettinger et al. (\citeproc{ref-jara-ettinger2022}{2022}) looked at shape bias in US adults across 3 shape exemplars and found a preference to extend nouns on the basis of shape rather than color or material.
The replication found an above-chance preference for shape, but with very high item-level variability.
The rescuer found that the first replication used stylized versions of the original stimuli taken from a different part of the paper.
The rescue used the same images of stimuli as the original, but still found strong item-level effects and relative preferences for color and material.

@@

Porter et al. (\citeproc{ref-porter2016}{2016}) looked at how linguistic intergroup bias influences judgments of shared group membership.
The first replication found an effect directionally inconsistent with the original result.
The rescuer obtained the direct wordings of the questions from the authors of the original paper, and noticed that they were substantially different from questions in the first replication.
The rescue used the same wordings as the original paper and replicated the effect, although the manipulation check failed.

Schechtman et al. (\citeproc{ref-schechtman2010}{2010}) looked at how widely people generalize tones that previously have been paired with losses or gains, and found that the loss-paired tones were generalized more.
The first replication found a statistically significant effect in the opposite direction of the original study.
It also found that people took much longer to learn which tones were associated with gains vs.~losses compared to the original.
The rescuer made several changes, including asking questions about the sound quality of the participant's computer.
The original study was conducted in person while the replication and rescue were online, so differences in speaker and headphone quality might have played a role in the different result.
The rescue found an effect in the same direction as the 1st replication (contrary to the original study).

@@

Yeshurun \& Levy (\citeproc{ref-yeshurun2003}{2003}) looked at the relationship between spatial attention and temporal resolution.
They measured the threshold where two flashes of light are perceived as one flash versus two as a function of whether the location of the flashes was pre-cued or not.
Yeshurun \& Levy (\citeproc{ref-yeshurun2003}{2003}) found that cueing led to less temporal resolution.
This was independently replicated by Chica \& Christie (\citeproc{ref-chica2009}{2009}), but was not replicated by either the student replication or the rescue.
Both the replication and rescue used in-person participants because the study relies on precise timing and details of the screen which are difficult to control online.
The first replication had a smaller sample size than the original, while the rescue matched the original study's sample size.
